---
title: "BWAS: Effect of categorical perturbation variables on c-Fos"
author: "Will Townes"
output: html_document
---

```{r}
#library(glmGamPoi)
library(cowplot)
library(ggplot2)
#library(ggridges)
library(MASS)
fp<-file.path
mkdir_p<-function(x){
    if(!dir.exists(x)){dir.create(x,recursive=TRUE)}
}
rp<-fp("results/bwas/perturbations")
mkdir_p(rp)
plt_dir<-fp("results/plots")
mkdir_p(plt_dir)
source("01_data_loading.R")
source("utils.R")
```

Analysis plan:
* Independent variables: within-batch categorical perturbations
* Outcomes: c-Fos for 122 brain regions
* Do linear regression with pcounts 
* Do negative binomial regression with raw counts and log of total count as offset using glm.nb from MASS package. If any warnings or errors, flag the region in the output.

## Main function

```{r}
#example of region prone to nb errors
batch="201904_ymaze_cfos"
conditions=c("CNO_control_no_reversal","CNO_control_reversal")
i<-61

#example of region prone to nb warnings
batch<-"201810_adultacutePC_ymaze_cfos"
conditions<-c("vector_control_reversal","lobVI_reversal")
i<-99
verbose<-FALSE

batch<-"202010_cfos"
conditions<-c("habituation","acquisition_day1")
i<-1

bwas<-function(batch,conditions,verbose=FALSE){
    #batch a single string
    #conditions a character vector of length two with control in first element,
    #treatment is in second element
    #rp the directory where the CSV files with results should be written
    ss<-cm$batch==batch & cm$condition %in% conditions
    cm2<-cm[ss,]
    cm2$condition<-factor(cm2$condition,levels=conditions)
    Yc<-counts122[ss,]
    Ypc<-pcounts122[ss,]
    nz<-colSums(Yc)>0
    Yc<-Yc[,nz]
    Ypc<-Ypc[,nz]
    lsz<-log(rowSums(Yc))
    #lsz<-lsz-mean(lsz)
    cnames<-colnames(Ypc)
    X<-with(cm2,model.matrix(~condition))
    #fit_gp<-glm_gp(t(Yc),X,offset=lsz)
    #thetas<-1/fit_gp$overdispersions
    bwas_lm<-function(i){
        fit<-lm(Ypc[,i]~X+0)
        res<-as.data.frame(summary(fit)$coef[2,,drop=FALSE])
        res$region<-cnames[i]
        res
    }
    bwas_nb<-function(i){
        if(verbose){print(i)}
        #ctl<-glm.control(maxit=100)
        fit<-tryCatch(withWarnings(glm.nb(Yc[,i]~X+0+offset(lsz))),
                      error=function(e){NULL})
        #fit is a list of $value, $warnings elements
        #$value contains the glm.nb fit object
        # fit<-glm(Yc[,i]~X+0, offset=lsz, fam=negative.binomial(theta=thetas[i]))
        if(is.null(fit) || !fit$value$converged){
            res<-matrix(NA,nrow=1,ncol=4)
            colnames(res)<-c("Estimate","Std. Error","z value","Pr(>|z|)")
            status<-"error"
            #return(data.frame(region=cnames[i],coef=NA,pval=NA))
        } else {
            res<-summary(fit$value,dispersion=1)$coef[2,,drop=FALSE]
            if(is.null(fit$warnings)){
                status<-"success"
            } else {
                status<-"warning"
            }
        }
        res<-as.data.frame(res)
        res$region<-cnames[i]
        res$status<-status
        res
    }
    idx<-seq_len(ncol(Ypc))
    res<-do.call(rbind,lapply(idx,bwas_lm))
    res$status<-"success"
    res2<-do.call(rbind,lapply(idx,bwas_nb))
    res$fdr_adj_pval<-p.adjust(res$`Pr(>|t|)`, "BH")
    res2$fdr_adj_pval<-p.adjust(res2$`Pr(>|z|)`, "BH")
    rownames(res2)<-rownames(res)<-NULL
    res2$control<-res$control<-conditions[1]
    res2$treatment<-res$treatment<-conditions[2]
    fname<-paste0(c(conditions,"pcounts-linreg.csv"),collapse="-")
    fname2<-paste0(c(conditions,"counts-nbreg.csv"),collapse="-")
    write.csv(res[,c(8,9,5,1:4,7,6)],fp(rp,fname),row.names=FALSE)
    write.csv(res2[,c(8,9,5,1:4,7,6)],fp(rp,fname2),row.names=FALSE)
}
```

## BWAS analyses

```{r}
# Acquisition Day 1 vs Habituation
bwas(batch="202010_cfos", conditions=c("habituation","acquisition_day1"))

# CNO reversal vs no reversal
bwas(batch="201904_ymaze_cfos", conditions=c("CNO_control_no_reversal","CNO_control_reversal"))

# CNO control vs vehicle control 
bwas(batch="202002_cfos", conditions=c("Vehicle_control_reversal","CNO_control_reversal"))

# Diluted vector control vs vehicle control
bwas(batch="202002_cfos", conditions=c("Vehicle_control_reversal","dilutedvector_control_reversal"))

# LobVI reversal vs CNO control
bwas(batch="201904_ymaze_cfos", conditions=c("CNO_control_reversal","lobVI_reversal"))

# LobVI reversal vs vector control
bwas(batch="201810_adultacutePC_ymaze_cfos", conditions=c("vector_control_reversal","lobVI_reversal"))
```

## Explore results

Find regions of interest
* relatively low counts
* strong contrast between conditions

```{r}
#batch<-"202010_cfos"
#conditions<-c("habituation","acquisition_day1")
batch<-"201904_ymaze_cfos"
conditions<-c("CNO_control_no_reversal","CNO_control_reversal")
fname2<-paste0(c(conditions,"counts-nbreg.csv"),collapse="-")
res<-read.csv(fp(rp,fname2),header=TRUE)
#subset to regions with FDR adj pval < .01
regs<-res$region[res$fdr_adj_pval<.01]
ss<-cm$batch==batch & cm$condition %in% conditions
cm2<-cm[ss,]
cm2$condition<-factor(cm2$condition,levels=conditions)
Yc<-counts122[ss,]
nz<-colSums(Yc)>0
Yc<-Yc[,nz]
lsz<-log(rowSums(Yc))
X<-with(cm2,model.matrix(~condition))
#list regions with at least one zero count
lc<-colnames(Yc)[apply(Yc,2,min)<10]
regs<-intersect(regs,lc)
```

Illustrate negative binomial fitting

```{r}
#iname<-"Nucleus raphe pontis"
#iname<-"Medial habenula"
#iname<-"Subparafascicular area"
#iname<-"Oculomotor nucleus"
#iname<-"Anterior tegmental nucleus"
#iname<-"Interanterodorsal nucleus of the thalamus"
plt_func<-function(iname,ret=c("plt","dat")){
    ret<-match.arg(ret)
    i<-which(colnames(Yc)==iname)
    fit<-glm.nb(Yc[,i]~X+0+offset(lsz))
    yhat<-predict(fit,type="response")
    lp<-dnbinom(Yc[,i],size=fit$theta,mu=yhat)
    mus<-tapply(yhat,cm2$condition,median)
    ranges<-tapply(Yc[,i],cm2$condition,range)
    pmf1<-function(x){dnbinom(round(x),fit$theta,mu=mus[1])}
    pmf2<-function(x){dnbinom(round(x),fit$theta,mu=mus[2])}
    pd<-data.frame(y=Yc[,i],cm2,pmf=lp)
    if(ret=="dat"){return(pd)}
    ggplot(pd)+geom_density(aes(x=y,fill=condition),color=NA,alpha=.8,adjust=2)+geom_function(fun=pmf1,n=1000,size=.8)+geom_function(fun=pmf2,n=1000,size=.8)+theme_bw()+ggtitle(iname)+xlab("c-Fos counts")+ylab("probability density")
}

plts<-lapply(regs,plt_func)
plot_grid(plotlist=plts)
#ggplot(pd)+stat_count(aes(x=y,fill=condition,y=..prop..),alpha=.9,position="identity",width=1)+geom_function(fun=pmf1,n=1000)+geom_function(fun=pmf2,n=1000)+theme_bw()+ggtitle(iname)+xlab("c-Fos counts")+ylab("probability density")
#+geom_point(aes(x=y,y=lp,color=condition))
#ggsave(fp(plt_dir,"nb_pmf_color.pdf"),width=6,height=4)
```

histogram instead of density

```{r}
iname<-"Anterodorsal nucleus"
i<-which(colnames(Yc)==iname)
fit<-glm.nb(Yc[,i]~X+0+offset(lsz))
yhat<-predict(fit,type="response")
lp<-dnbinom(Yc[,i],size=fit$theta,mu=yhat)
mus<-tapply(yhat,cm2$condition,median)
ranges<-tapply(Yc[,i],cm2$condition,range)
pmf1<-function(x){dnbinom(round(x),fit$theta,mu=mus[1])}
pmf2<-function(x){dnbinom(round(x),fit$theta,mu=mus[2])}
pd<-data.frame(y=Yc[,i],cm2,pmf=lp)
ggplot(pd)+geom_histogram(aes(x=y,fill=condition,y=..density..),color=NA,alpha=.8,position="identity",bins=6,boundary=0)+geom_function(fun=pmf1,n=1000,size=.8)+geom_function(fun=pmf2,n=1000,size=.8)+theme_bw()+ggtitle(iname)+xlab("c-Fos counts")+ylab("probability density")
ggsave(fp(plt_dir,"nb_pmf_anterodorsal.pdf"),width=6,height=4)

iname<-"Interanteromedial nucleus of the thalamus"
i<-which(colnames(Yc)==iname)
fit<-glm.nb(Yc[,i]~X+0+offset(lsz))
yhat<-predict(fit,type="response")
lp<-dnbinom(Yc[,i],size=fit$theta,mu=yhat)
mus<-tapply(yhat,cm2$condition,median)
ranges<-tapply(Yc[,i],cm2$condition,range)
pmf1<-function(x){dnbinom(round(x),fit$theta,mu=mus[1])}
pmf2<-function(x){dnbinom(round(x),fit$theta,mu=mus[2])}
pd<-data.frame(y=Yc[,i],cm2,pmf=lp)
ggplot(pd)+geom_histogram(aes(x=y,fill=condition,y=..density..),color=NA,alpha=.8,position="identity",bins=9,boundary=0)+geom_function(fun=pmf1,n=1000,size=.8)+geom_function(fun=pmf2,n=1000,size=.8)+theme_bw()+ggtitle(iname)+xlab("c-Fos counts")+ylab("probability density")
ggsave(fp(plt_dir,"nb_pmf_interanteromedial.pdf"),width=6,height=4)
```

```{r}
fnames<-list.files(fp(rp))
readfile_func<-function(fname){
    res<-cbind(filename=fname,read.csv(fp(rp,fname),header=TRUE))
    colnames(res)[7:8]<-c("effect_size","pval")
    res
}
res<-do.call(rbind,lapply(fnames,readfile_func))
res$mod<-sapply(strsplit(res$filename,"-"),function(x){sub(".csv","",x[[4]],fixed=TRUE)})
with(subset(res,mod=="nbreg"),table(status))
bad<-subset(res,mod=="nbreg" & status!="success")
extreme<-subset(res,pval<1e-16 & status=="success")
write.csv(bad,"results/bwas/perturbations_bad.csv",row.names=FALSE)
write.csv(extreme,"results/bwas/perturbations_extreme.csv",row.names=FALSE)
```

Make sure interesting neocortical regions not included

```{r}
regs<-c("Infralimbic area","Anterior cingulate area","Prelimbic area","Temporal association areas","Somatosensory areas")
any(regs %in% bad$region)
```

Regions with extreme p-values, is it a real difference?

```{r}
batch<-"201904_ymaze_cfos"
conditions<-c("CNO_control_reversal","lobVI_reversal")
region<-"Posterior complex of the thalamus"
ss<-cm$batch==batch & cm$condition %in% conditions
Yp<-pcounts122[ss,]
Yc<-counts122[ss,]
x<-factor(cm$condition[ss])
boxplot(Yc[,region]~x,xlab="experimental condition",ylab="counts",main="Posterior complex of the thalamus",log="y")
boxplot(Yp[,region]~x,xlab="experimental condition",ylab="pcounts",main="Posterior complex of the thalamus")
lsz<-log(rowSums(Yc))
fit_gp<-glm_gp(t(Yc),model.matrix(~x),offset=lsz)
thetas<-1/fit_gp$overdispersions
fit<-glm.nb(Yc[,region]~x+offset(lsz))
fit2<-glm(Yc[,region]~x+offset(lsz),fam=negative.binomial(theta=thetas[region]))
```

Double check data for weird outliers:
lobVI_reversal vs vector_control_reversal
* Pontine reticular nucleus
* Pedunculopontine nucleus
Flagged numerical warning in nb regression for both regions.

```{r}
batch<-"201810_adultacutePC_ymaze_cfos"
conditions<-c("vector_control_reversal","lobVI_reversal")
region<-"Pontine reticular nucleus"
#region<-"Pedunculopontine nucleus"
ss<-cm$batch==batch & cm$condition %in% conditions
Yp<-pcounts122[ss,]
Yc<-counts122[ss,]
x<-factor(cm$condition[ss])
boxplot(Yc[,region]~x,xlab="experimental condition",ylab="counts",main=region)
boxplot(Yp[,region]~x,xlab="experimental condition",ylab="pcounts",main=region)
lsz<-log(rowSums(Yc))
#fit_gp<-glm_gp(t(Yc),model.matrix(~x),offset=lsz)
#thetas<-1/fit_gp$overdispersions
summary(fit<-glm.nb(Yc[,region]~x+offset(lsz)))$coefficients
summary(fit2<-glm(Yc[,region]~x+offset(lsz),fam=negative.binomial(theta=thetas[region])),dispersion=1)$coefficients
summary(lm(Yp[,region]~x))
fit$converged
```

Vehicle_control_reversal
202002_cfos
Nucleus incertus
Flagged numerical warning in nb regression.

```{r}
batch<-"202002_cfos"
conditions<-c("Vehicle_control_reversal","CNO_control_reversal")
region<-"Nucleus incertus"
ss<-cm$batch==batch & cm$condition %in% conditions
Yp<-pcounts122[ss,]
Yc<-counts122[ss,]
x<-factor(cm$condition[ss])
boxplot(Yc[,region]~x,xlab="experimental condition",ylab="counts",main=region)
boxplot(Yp[,region]~x,xlab="experimental condition",ylab="pcounts",main=region)
lsz<-log(rowSums(Yc))
fit_gp<-glm_gp(t(Yc),model.matrix(~x),offset=lsz)
thetas<-1/fit_gp$overdispersions
summary(fit<-glm.nb(Yc[,region]~x+offset(lsz)))$coefficients
summary(fit2<-glm(Yc[,region]~x+offset(lsz),fam=negative.binomial(theta=thetas[region])),dispersion=1)$coefficients
summary(lm(Yp[,region]~x))
fit$converged
```

Tuberal nucleus
Numerical warning in nb regression flagged but result consistent with linear regression (p-value exaggerated however).

```{r}
batch<-"202002_cfos"
conditions<-c("Vehicle_control_reversal","dilutedvector_control_reversal")
region<-"Tuberal nucleus"
ss<-cm$batch==batch & cm$condition %in% conditions
Yp<-pcounts122[ss,]
Yc<-counts122[ss,]
x<-factor(cm$condition[ss])
boxplot(Yc[,region]~x,xlab="experimental condition",ylab="counts",main=region)
boxplot(Yp[,region]~x,xlab="experimental condition",ylab="pcounts",main=region)
lsz<-log(rowSums(Yc))
fit_gp<-glm_gp(t(Yc),model.matrix(~x),offset=lsz)
thetas<-1/fit_gp$overdispersions
summary(fit<-glm.nb(Yc[,region]~x+offset(lsz)))$coefficients
summary(fit2<-glm(Yc[,region]~x+offset(lsz),fam=negative.binomial(theta=thetas[region])),dispersion=1)$coefficients
summary(lm(Yp[,region]~x))
fit$converged
```

Periaqueductal gray and lateral habenula boxplots

```{r}
od<-fp(plt_dir,"boxplots")
mkdir_p(od)
regs<-c("Periaqueductal gray","Lateral habenula")
Yp<-pcounts122[,regs]
conditions<-data.frame(
    batch=c("202010_cfos", "201904_ymaze_cfos", "201904_ymaze_cfos"),
    control=c("habituation", "CNO_control_no_reversal", "CNO_control_reversal"),
    treatment=c("acquisition_day1","CNO_control_reversal","lobVI_reversal"))
for(i in 1:nrow(conditions)){
    for(reg in regs){
        batch<-conditions$batch[i]
        cds<-conditions[i,2:3]
        ss<-cm$batch==batch & cm$condition %in% cds
        pd<-data.frame(condition=factor(cm$condition[ss],levels=cds), pcounts=pcounts122[ss,reg])
        ggplot(pd,aes(x=condition,y=pcounts))+geom_boxplot(outlier.shape=NA,fill="gray")+geom_jitter(height=0,width=.1,size=3)+ylab("proportion of counts")+ggtitle(reg)
        fname<-paste0(paste(c(reg,cds),collapse="-"),".pdf")
        ggsave(fp(od,fname),width=6,height=4)    
    }
}
```

## Aggregated regions

```{r}
agg<-read.csv(fp("data/original/aggregated_regions.csv"),header=TRUE)
xt<-as.matrix(table(agg$region,agg$aggregated_region))
Y<-counts122
lsz0<-log(rowSums(Y))
Y<-Y[,rownames(xt)]
Y2<-Y%*%xt

bwas2<-function(batch,conditions,verbose=FALSE){
    #batch a single string
    #conditions a character vector of length two with control in first element,
    #treatment is in second element
    #rp the directory where the CSV files with results should be written
    ss<-cm$batch==batch & cm$condition %in% conditions
    cm2<-cm[ss,]
    cm2$condition<-factor(cm2$condition,levels=conditions)
    Yc<-Y2[ss,]
    lsz<-lsz0[ss]
    cnames<-colnames(Yc)
    X<-with(cm2,model.matrix(~condition))
    bwas_nb<-function(i){
        if(verbose){print(i)}
        #ctl<-glm.control(maxit=100)
        fit<-tryCatch(withWarnings(glm.nb(Yc[,i]~X+0+offset(lsz))),
                      error=function(e){NULL})
        #fit is a list of $value, $warnings elements
        #$value contains the glm.nb fit object
        if(is.null(fit) || !fit$value$converged){
            res<-matrix(NA,nrow=1,ncol=4)
            colnames(res)<-c("Estimate","Std. Error","z value","Pr(>|z|)")
            status<-"error"
        } else {
            res<-summary(fit$value,dispersion=1)$coef[2,,drop=FALSE]
            if(is.null(fit$warnings)){
                status<-"success"
            } else {
                status<-"warning"
            }
        }
        res<-as.data.frame(res)
        res$region<-cnames[i]
        res$status<-status
        res
    }
    idx<-seq_len(ncol(Yc))
    res<-do.call(rbind,lapply(idx,bwas_nb))
    res$fdr_adj_pval<-p.adjust(res$`Pr(>|z|)`, "BH")
    rownames(res)<-NULL
    res$control<-conditions[1]
    res$treatment<-conditions[2]
    #fname2<-paste0(c(conditions,"counts-nbreg.csv"),collapse="-")
    #write.csv(res[,c(8,9,5,1:4,7,6)],fp(rp,fname2),row.names=FALSE)
    res[,c(8,9,5,1:4,7,6)]
}
res<-list()
res[[1]]<-bwas2(batch="202010_cfos", conditions=c("habituation","acquisition_day1"))

# CNO reversal vs no reversal
res[[2]]<-bwas2(batch="201904_ymaze_cfos", conditions=c("CNO_control_no_reversal","CNO_control_reversal"))

# LobVI reversal vs CNO control
res[[3]]<-bwas2(batch="201904_ymaze_cfos", conditions=c("CNO_control_reversal","lobVI_reversal"))
res<-do.call(rbind,res)
write.csv(res,"results/bwas/perturbations_aggregated.csv",row.names=FALSE)
```
